%!TEX root = ../main.tex

\chapter{Ribofinder}
\label{ch:rfinder}

\lhead{Ribofinder: A \Rb Detection Pipeline}

\section{Introduction}
\label{sec:rfinder:intro}

In this chapter, we present the \rfinder program---a pipeline to facilitate the
detection of putative guanine \rbs across genomic data. The \rfinder
tool operates in three stages. First we use \infernal and \tthp to detect
putative aptamers and expression platforms, two distinct components of
\rbs described in section \ref{sec:rfinder:bkgrnd}. After coalescing
this data into a pool of candidate \rbs, we use \rfold with constraints
based on experimental data to compute the two distinct structural
conformations---`gene on' and `gene off'. In the third and final stage, we
leverage \foldalign to measure the similarity between our candidate pool and a
canonical guanine \rb well studied in the literature, the xpt G-box
\rb from Bacillus subtilis.

\subsection{Organization}
\label{subsec:rfinder:org}

This chapter is organized in the following fashion. After providing background
on the structural components of a \rb alongside their biological
significance, we outline the deficiencies in the `state of the art' software
when as it relates specifically to \rb detection. We then move on to outline
the three stages of \rfinder: candidate selection, structural prediction, and
candidate curation. Having described the approach of the software, we move on
to present our findings in using \rfinder to detect guanine \rbs across
the bacterial RefSeq database. Finally, we provide brief commentary on possible
extensions of the algorithm to locate other flavors of \rbs, of which
adenine-sensitive aptamers are a straightforward extension.

\section{Background}
\label{sec:rfinder:bkgrnd}

\section{The \rfinder pipeline}
\label{sec:rfinder:pipeline}

\subsection{Step 1: Candidate selection}
\label{subsec:rfinder:selection}

\subsection{Step 2: Structural prediction}
\label{subsec:rfinder:consfold}

\subsection{Step 3: Candidate curation}
\label{subsec:rfinder:curation}

\section{Using \rfinder against the RefSeq database}
\label{sec:rfinder:refseq}

\section{Extending beyond guanine \rbs}
\label{sec:rfinder:ext}
